---
title: "Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4, fig.align = "center")
```

This tutorial learns how to use created sentiment measures in both a high-dimensional regression context proposed by the package, as well as in any other (regression) model.

**Preparation**
&nbsp;

```{r}
library("sentometrics")
library("ggplot2")
library("data.table")
library("zoo")
library("e1071")

data("usnews")
data("epu")
data("list_lexicons")
data("list_valence_shifters")

set.seed(505)
```

We create several sentiment time series to use in the models.

```{r}
corpus <- sento_corpus(usnews)
lexicons <- sento_lexicons(list_lexicons[c("GI_en", "HENRY_en")])
ctr <- ctr_agg("counts", "proportional", c("linear", "equal_weight"), by = "month", lag = 6)
measures <- sento_measures(corpus, lexicons, ctr)
```

As target variable, we take the built-in Economic Policy Uncertainty (EPU) index, and randomly generate two other covariates.

```{r}
y <- epu[epu$date %in% get_dates(measures), "index"]
x <- data.frame(x1 = runif(length(y)), x2 = rnorm(length(y))) # two other (random) x variables

length(y) == nobs(measures)
```

**Background**
&nbsp;

The model supported is the elastic net regression, made available as a finetuned interface to the `glmnet::glmnet()` function. The elastic net has two hyperparameters which need to be calibrated: alpha (where if 1 one gets the LASSO regression, and if 0 one gets the Ridge regression), and lambda (which defines the degree of regularization, where if 0 there will be none). The possible options to calibrate these parameters are to use an information criterion, or to use time series cross-validation.

### In-sample regression

The regression specifications are defined through the `ctr_model()` function. A `"gaussian"` model means a linear regression. The `type` argument decides on the hyperparameter calibration method, in this case, a Bayesian information criterion will be used. The model is run with `sento_model()`, and results in a `sento_model` object.

```{r}
ctrIC <- ctr_model(model = "gaussian", type = "BIC") # same as ctr_model()
out1 <- sento_model(measures, y, x, ctr = ctrIC)

summary(out1)
```

To do predictions with the model, there is the `predict()` function.

```{r}
nx <- nmeasures(measures) + ncol(x)
newx <- runif(nx) * cbind(as.data.table(measures)[, -1], x)[30:39, ] # 10 random new predictors

predict(out1, newx = as.matrix(newx))
```

The `ctr_model()` function provides much more flexibility. Below setup uses a cross-validation calibration approach, for a rolling in-sample size of 70 and an out-of-sample test set of 10. This is typically more time-consuming, hence, we show how one can parallelize the single model call. We also take the differences of the target variable, and shift it four periods ahead. The data reorganization is handled internally.

```{r}
library("parallel")
library("doParallel")

cl <- makeCluster(2) # two cores
registerDoParallel(cl)

ctrCV <- ctr_model(model = "gaussian", type = "cv", h = 4, do.difference = TRUE, trainWindow = 70,
                   testWindow = 10, alphas = c(0.10, 0.50, 0.90), do.progress = FALSE)
out2 <- sento_model(measures, y, x = x, ctr = ctrCV)

stopCluster(cl)
registerDoSEQ() # ensures sequential processing hereafter

summary(out2)
```

The `sento_model()` function supports both **linear** and **logistic** (binomial and multinomial) regression. It suffices to change the `model` argument in `ctr_model()` to switch between models, assuming you provide the right target input data.

### Iterative regressions and out-of-sample prediction

Setting `do.iter = TRUE` and picking a `nSample` size will run regressions iteratively moving forward over time. At every run, as in a real forecasting experiment, an out-of-sample prediction is generated and stored. To speed up the analysis, one could try to augment the `nCore` argument which will enact parallel computation across the different iterations.

```{r}
ctrIter <- ctr_model(model = "gaussian", type = "AIC", do.iter = TRUE, h = 3,
                     oos = 2, alphas = c(0.25, 0.75), nSample = 75, do.progress = FALSE)
out3 <- sento_model(measures, y, x = x, ctr = ctrIter)

summary(out3)
```

The out-of-sample predictions can be plot against the true realizations of the target variable.

```{r}
plot(out3)
```

### Attribution analysis

Take the model output from the previous example. To analyze the contribution of each sentiment variable component to a prediction, called attribution, the `attributions()` function does the trick.

```{r}
attr3 <- attributions(out3, measures, refDates = names(out3$models)[1:50])
```

```{r}
plot(attr3, "lexicons")
plot(attr3, "features")
plot(attr3, "time")
```

For a very deep analysis, it is also possible to trace back the contribution of a single document to a prediction at a given point in time. As the time lag for the sentiment measures was set to 6 months, news articles up to 5 months before a prediction date have an impact, the degree depending on the sentiment and the associated time weight.

```{r}
attr3$documents$`2004-06-01`
```

### Other models

The `sento_model()` function is no perfect toolbox. However, by converting the sentiment measures into a straight `data.table` (or a `data.frame`), you are not limited. For instance, below example shows how to do a low-dimensional regression using the `lm()` function, with some of the sentiment variables created in the package.

```{r}
dt <- as.data.table(measures)
z <- zoo(dt[, 2:7], order.by = dt[["date"]])
reg <- lm(y ~ z)
```

At some point we might integrate other useful models into the `sento_model()` function, but having the sentiment measures at your disposal, you are already good to go elsewhere.

### Classification using sentiment

The above examples take a time series perspective. However, sentiment on text-level can also be useful to predict other attributes of a text. We illustrate one possible approach hereunder, where we try to predict if news discusses the economy using sentiment information.

```{r}
y <- as.factor(usnews$economy)
```

To calculate a couple of sentiment variables, we split the Harvard General Inquirer lexicon into 15 random groups.

```{r}
l <- list_lexicons$GI_en
lg <- split(l, factor(round(14 * runif(nrow(l))), labels = paste0("s", 1:15)))
lexicons <- sento_lexicons(lg)

nrow(l) == nrow(rbindlist(lexicons))
```

Sentiment is computed in one go using the `compute_sentiment()` function, this time using an inverse U-shaped scheme for normalizing the scores.

```{r}
s <- compute_sentiment(usnews$texts, lexicons, "inverseUShaped")

colnames(s)
```

The data is divided into a training set, and a test set for model validation.

```{r}
data <- cbind(y = y, s[, -c(1:2)])
dataTrain <- data[1:3500, ]
dataTest <- data[3501:nrow(data), ]
```

We employ support vector machine classification to estimate our model. 

```{r}
model <- svm(y ~ ., dataTrain)
```

Let's construct confusion matrices to see how well the model does in-sample, and in the test set.

```{r}
conf_matrix <- function(m, d) table(predict(m, d), d$y, dnn = c("Prediction", "Actual"))
```

```{r}
conf_matrix(model, dataTrain)
conf_matrix(model, dataTest)
```

Not good at all, none of the economic articles are picked up out-of-sample! Off-the-shelf modeling is rarely a good idea. Time to go back to the drawing table and better tune those sentiment predictors and model parameters. Maybe the random lexicon splitting was not such a good idea?

